---
title: "Functions Demo"
author: "Rachael Uhlmann"
date: "2026-02-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/rachaelsmacbook/Desktop/PAF514 Final ")

library(dplyr)
library(pander)
library(knitr)
library(gender)
source("utils.R")
```

```{r}
URL <- "https://docs.google.com/spreadsheets/d/1RoiO9bfpbXowprWdZrgtYXG9_WuK3NFemwlvDGdym7E/export?gid=1335284952&format=csv"
d <- read.csv(URL)
```
## Step 2: Parse First Name

The get_first_name() function extracts first names from the Full.Name field 
which is formatted as "LAST, FIRST MIDDLE". It splits on the comma, trims 
whitespace, and returns only the first word to handle compound names and 
middle names consistently across all years of data.

```{r}

get_first_name <- function(name)
{
  # split on comma
  parts <- strsplit(name, ",")
  
  # take what comes after the comma
  first_part <- sapply(parts, function(x){ x[2] })
  
  # trim whitespace
  first_part <- trimws(first_part)
  
  # take only first word (handles middle names)
  first_part <- sapply(strsplit(first_part, " "), function(x){ x[1] })
  
  return(first_part)
}
  
```

```{r}

# test get_first_name
test_names <- head(d$Full.Name, 25)
first_names <- get_first_name(name = test_names)
data.frame(test_names, first_names)
```
## Step 3: Add Gender

The add_gender() function queries the gender package using US birth 
certificate data to assign probable gender based on first name. Names 
not found in the database are coded as "uncoded" rather than left as NA.

```{r}
 
# Step 5: Merge gender back to dataset 

add_gender <- function(d) 
{ # Step 1: Extract first names
 d$first.name <- get_first_name(d$Full.Name) 
# Step 2: Get unique first names only 
unique.names <- unique(d$first.name) 
# Step 3: Query gender package 
gen <- gender(unique.names) 
# Step 4: Keep only name and gender columns 
gen <- gen %>% select(name, gender) 
# Step 5: Merge gender back to dataset 
d <- merge(d, gen, 
by.x = "first.name", 
by.y = "name", 
all.x = TRUE) 
# Step 6: Recode NAs as uncoded 
d$gender[is.na(d$gender)] <- "uncoded" 
# Step 7: Make gender a factor 
d$gender <- factor(d$gender, 
levels=c("male","female","uncoded")) 
return(d) }
 
```
 
```{r}
library(genderdata)
library(gender)

# Test gender package directly
test <- gender(c("James", "Mary", "John"))
test
```

## Step 4: Standardize Job Titles

The add_titles() function recodes the raw Job.Description field into five standardized categories: Full Professor, Associate Professor, Assistant Professor, Teaching Faculty, and Researcher. This standardization is necessary because job titles vary inconsistently across departments and years in the raw data.

```{r}
# Check what your first names look like
head(unique(d$first.name), 20)

```
```{r}
# Test add_gender
d <- add_gender(d)

# Check results
head(d$gender)
table(d$gender)
```
```{r}
# Reload fresh data
URL <- "https://docs.google.com/spreadsheets/d/1RoiO9bfpbXowprWdZrgtYXG9_WuK3NFemwlvDGdym7E/export?gid=1335284952&format=csv"
d <- read.csv(URL)

# Now run add_gender on fresh data
d <- add_gender(d)

# Check results
head(d$gender)
table(d$gender)
```
```{r}
# Test add_titles
d <- add_titles(d)

# Check results
table(d$title)
```
## Steps 5 & 6: Convert and Standardize Salary

The fix_salary() function combines two steps — converting salary strings to numeric values by removing dollar signs and commas, then adjusting all salaries to full-time equivalent (FTE) so part-time and full-time employees can be compared fairly. Outliers include coaching staff and hourly workers 
who are removed by get_study_sample() in Step 7.


```{r}
# Test fix_salary
d <- fix_salary(d)

# Check results
summary(d$salary)
head(d$salary)
```
```{r}
# Find the $3.5 million salary
d %>% 
  filter(salary > 500000) %>%
  select(Full.Name, Job.Description, Salary, FTE, salary)
# Although these salaries are extremely high, this correlates to the salaries generally paid to coaches and presidents as well as other high positions. 

```
```{r}
# Find suspiciously low salaries
d %>%
  filter(salary < 100) %>%
  select(Full.Name, Job.Description, Salary, FTE, salary)
# These are most definitely hourly rates for accompanists. That is why this number looks so strange.
```
## Step 7: Select Study Sample

The get_study_sample() function filters the full ASU salary dataset down 
to only academic faculty in approved academic units. This removes coaches, 
administrators, and other non-faculty roles that would skew salary 
comparisons. After filtering, 2,459 employees remain from the original 
12,520.



```{r}

# Test get_study_sample
d <- get_study_sample(d)

# How many rows after filtering?
nrow(d)

# Verify titles look correct
table(d$title)
```


```{r}
create_salary_table <- function(d)
{
  t.salary <-
    d %>%
    filter(!is.na(title) & title != "") %>%
    group_by(title, gender) %>%
    summarize(q25 = quantile(salary, 0.25),
              q50 = quantile(salary, 0.50),
              q75 = quantile(salary, 0.75),
              n = n()) %>%
    ungroup() %>%
    mutate(p = round(n/sum(n), 2))
  
  return(t.salary)
}

```

## Step 8: Summarize Salaries by Rank and Gender

The create_salary_table() function groups faculty by title and gender, 
calculating the 25th, 50th, and 75th percentile salaries for each group. 
Results reveal consistent gender pay gaps at every faculty rank, with male 
faculty earning higher median salaries than female faculty across all five 
title categories.

```{r}
# Test create_salary_table
t.salary <- create_salary_table(d)
pander(t.salary)
```

```{r}
get_top5 <- function(d)
{
  d %>%
    arrange(desc(salary)) %>%
    select(Full.Name, gender, Job.Description, salary) %>%
    mutate(salary = dollarize(salary)) %>%
    head(5)
}
```

```{r}
# Test get_top5
get_top5(d)
```
## Step 10: Top Five Salaries

The get_top5() function returns the five highest paid employees in a given department, displaying name, gender, job title, and formatted salary. This provides quick context for understanding the salary ceiling within each academic unit.

---

## Part II: Department Loop

The following loop iterates over every unique department in the study sample and prints a formatted salary summary table and top 5 earners for each department. This allows us to compare gender pay patterns across all academic units in a single automated pass.

```{r department-loop, results='asis'}

# Get sorted list of all unique departments in the study sample
departments <- sort(unique(d$Department.Description))

cat("**Total departments in study sample:**", length(departments), "\n\n")

for (dept in departments) {
  
  # Filter to current department
  d.dept <- d %>% filter(Department.Description == dept)
  
  # Skip departments with fewer than 5 employees (too small for meaningful comparison)
  if (nrow(d.dept) < 5) next
  
  # Print department header
  cat("\n\n###", dept, "\n\n")
  cat("**N =**", nrow(d.dept), "faculty  |  ")
  cat("**Gender breakdown:** Male =", sum(d.dept$gender == "male", na.rm = TRUE), 
      ", Female =", sum(d.dept$gender == "female", na.rm = TRUE), 
      ", Uncoded =", sum(d.dept$gender == "uncoded", na.rm = TRUE), "\n\n")
  
  # Salary table by title and gender
  t.dept <- create_salary_table(d.dept)
  
  if (nrow(t.dept) > 0) {
    cat("**Salary Summary by Title and Gender:**\n\n")
    print(kable(t.dept, 
                format = "html",
                col.names = c("Title", "Gender", "25th %ile", "Median", "75th %ile", "N", "Proportion"),
                digits = 0,
                format.args = list(big.mark = ",")))
    cat("\n\n")
  }
  
  # Top 5 earners in department
  cat("**Top 5 Earners:**\n\n")
  top5.dept <- get_top5(d.dept)
  print(kable(top5.dept, format = "html"))
  cat("\n\n---\n\n")
}
```

